prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis

### 2022-10-21 inst/MyArticle/paper.Rmd
using figure1 figure2 figure3

---
title: prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis
author: 
  - Witold E. Wolski $^{\dagger\ddagger}$^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - [University of Zurich](https://uzh.ch)/[ETH Zurich](https://ethz.ch), Winterthurerstrasse 190, CH-8057 Zurich, Switzerland. $^\ddagger$[Swiss Institute of Bioinformatics (SIB), Quartier Sorge - Batiment Amphipole, 1015 Lausanne, Switzerland](https://www.sib.swiss/) $^\dagger$Correspondence&colon; \texttt{wew@fgcz.ethz.ch}]
  - Paolo Nanni$^{\ast}$
  - Jonas Grossmann$^{\ast\ddagger}$
  - Maria d'Errico$^{\ast\ddagger}$
  - Ralph Schlapbach$^{\ast}$
  - Christian Panse$^{\ast\ddagger}$
output: 
  bookdown::pdf_document2:
      toc: true
      toc_appendix: false
      toc_bib: false
      keep_tex: true
date: "`r format(Sys.time())`"
bibliography: prolfqua.bib
vignette: >
  %\VignetteIndexEntry{prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
abstract: |
    Mass spectrometry is widely used for quantitative proteomics studies, relative protein quantification, and differential expression analysis of proteins. There is a large variety of signal processing software and analysis tools. Nevertheless, there is a need for a modular, easy-to-use application programming interface in R that transparently supports a variety of well principled statistical procedures to make applying them to proteomics data, comparing and understanding their differences easy. The `r BiocStyle::Githubpkg("fgcz/prolfqua")` package integrates essential steps of the mass spectrometry-based differential expression analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. The package makes integrating new data formats easy. It can be used to model simple experimental designs with a single explanatory variable and complex experiments with multiple factors and hypothesis testing. The implemented methods allow sensitive and specific differential expression analysis. Furthermore, the package implements benchmark functionality that can help to compare data acquisition, data preprocessing, or data modeling methods using a gold standard dataset. The application programmer interface of `r BiocStyle::Githubpkg("fgcz/prolfqua")`  strives to be clear,  predictable, discoverable, and consistent to make proteomics data analysis easy and exciting. Finally, the `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package is available on GitHub https://github.com/fgcz/prolfqua, distributed under the MIT license. It runs on all platforms supported by the [R free software environment for statistical computing and graphics](https://www.r-project.org/).

@@

```{r contrast2, eval=TRUE, echo=TRUE}
mlm <- prolfqua::strategy_lm(
  transformedIntensity ~ condition_ + batch_ + condition_:batch_
  )
Yeast2Factor <- prolfqua::prolfqua_data("data_Yeast2Factor")
mm <- prolfqua::build_model(Yeast2Factor$data, mlm, subject_Id = "protein_Id")

@@

The code snippets in this section demonstrate how a differential expression analysis workflow can be implemented using the `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package. To speed up the computation of these examples, we use a subset of the Ionstar dataset generated by randomly selecting $400$ proteins. First, we remove all proteins with a single peptide and all observations for which MaxQuant reports zero intensities which leaves $332$ proteins. Next, peptide abundances are $\log_2$ transformed and robust z-score scaled using the method `robscale`. Then, using the `LFQDataPlotter` class, we show the distribution of the normalized peptide abundances in Figure \@ref(fig:prepro) Panel A. Afterwards, protein intensities are estimated from peptide intensities using Tukey's median polish. Figure \@ref(fig:prepro) Panel B shows the peptide intensities and the estimated protein intensities. Next, we compute the standard deviation of all the proteins in each group and display their distribution using violin plots (Panel C). Finally, we create a boxplot (Panel D) showing the abundance of one protein.

```{r prepro,  echo = TRUE , fig.cap="(ref:scaling)", out.width = '90%', fig.width = 8, fig.height = 8}
R.version.string; packageVersion("prolfqua")

@@

The following code example illustrates how we compute differences among groups. First, the linear model and the differences are specified. Afterward, the model is fitted to the data using the `build_model` function. Next, we estimate the contrasts either from the linear model using the `Contrasts` class or directly from the data using the `ContrastsSimpleImpute` class. Afterward, we apply $t$-statistic moderation using the `ContrastModerated` class. Finally, the `merge_contrasts_results` function merges both sets of contrast estimates, preferring the one obtained from the linear model if both are available. Then we create the plots shown in Figure \@ref(fig:exampleContrasts). Panel A shows the distribution of the $p$-values, Panel B shows the volcano plot for each comparison, and Panel C shows a Bland-Altman plot which shows the difference between groups as a function of the rank of the protein abundance.

(ref:exampleContrasts) Panel A - Histogram showing the distribution of p-values for $332$ proteins. Panel B -  Volcano plot showing $-\log_{10}$ transformed $FDR$ as a function of the difference between groups for $332$ proteins. With black dots, we show effect size and FDR estimates obtained from the linear model, while in green, we plot those obtained using imputation. Finally, panel C shows the difference between groups as a function of the rank of the abundance of the proteins.

```{r exampleContrasts, fig.cap="(ref:exampleContrasts)", echo=TRUE, out.width = '90%', fig.width = 8, fig.height = 5}
# specify differences among groups
contrasts <- c(
  "dilution_(9/7.5)_1.2" =   "dilution.e - dilution.d",
  "dilution_(7.5/6)_1.25" =   "dilution.d - dilution.c"
)
## fit model
lmmodel <- paste(lfqpro$response()," ~ dilution.")
modelFunction <- prolfqua::strategy_lm( lmmodel, model_name = "lm")
models <- prolfqua::build_model(lfqpro, modelFunction)

@@

When comparing differential expression analysis performance, a relevant parameter is the number of proteins for which a method estimated differences (see Figure \@ref(fig:FDRfdp) Panel A), which indicates how robust the procedure works in the presence of missing observations. For each protein, we tried to determine four differences ($\Delta = (1.20, 1.25, 1.30, 1.50)$). Given 4046 proteins with at least two peptides, there are in total $16184$ possible differences. However, some methods can not estimate all of them. The set of proteins with effect size estimates might differ for each method, biasing direct comparison of scores such as the $pAUC$. For instance, we observe that `r BiocStyle::Biocpkg("MSstats")` estimates $16058$ group differences while the mixed effect models estimates the fewest with $15940$. Hence, to conclude that one method shows a better performance, it does not suffice if the $pAUC$ is greater, but the number of proteins with differential expression results needs to be equal or larger.

Figure \@ref(fig:FDRfdp) Panel B shows how various estimates obtained from the models, i.e., the difference between groups, $t$-statistics, and scaled $p$-values allow identifying true positives (TP) given a false positive rate (FPR) of $10%$, by displaying the partial area under the ROC. Ordering proteins by the $t$-statistic or $p$-value leads to a higher $pAUC_{10}$ than when ordering by the estimated difference among groups. Suppose an accurate estimate of the difference among groups is essential. In that case, the linear models fitted to protein intensities, calculated using Tukey's median polish, perform better than those directly modeling peptide intensities, e.g., `ropeca` or `prot_mixed_effect_moderated` (see Table \@ref(tab:describeBenchmarkedModels) Abundance column). We speculate that outliers at the peptide level do not affect the protein estimates when using Tukey's median polish method, a non-parametric method to infer protein abundances. This hypothesis could be examined, by including other forms of protein intensity inferences implemented in `prolfqua`, e.g. `top-N` or `rlm`, into the benchmark.

There are only minor differences in the $pAUC_{10}$ between the $t$-statistics or the scaled $p$-value  (see Figure \@ref(fig:FDRfdp) Panel A). Interestingly, the $pAUC$ increases when using $p$-values instead of the $t$-statistics for linear models, while it decreases for mixed effect models. The reason is an erroneous denominator degree of freedom estimation for many proteins, a known problem in the case of mixed effect models.

We also benchmark if the $FDR$ obtained from a model is an unbiased estimate of the false discovery proportion $FDP$. Figure \@ref(fig:FDRfdp) Panel C shows the $FDP$, obtained from the confusion matrix, as a function of the $FDR$ determined from the model. Most lines are below the diagonal, indicating that the $FDR$ estimates are modestly conservative for this particular benchmark dataset. In the case of `r BiocStyle::Biocpkg("MSstats")`, we observe a high proportion of false discoveries for small $FDR$ values. In the case of the ROPECA method, the $FDR$ estimates, obtained by applying the Benjamini-Hochberg correction to the Beta distribution-based regulation probabilities, strongly overestimate the $FDP$.

(ref:FDRfdp) Panel A - Number of estimated contrasts for each modeling method (higher is better). Panel B - Partial area under the ROC curve at $10\%$ FPR ($pAUC_{10}$) for all contrasts and three different statistics: the difference among groups (diff - Panel B left), the scaled $p$-value (sign(diff) $\cdot$ p.value) (scaled.p.value - Panel B center) and the $t$-statistics (statistic - Panel B right), where a higher pAUC is better. The red line indicates the average area under the curve of all methods.  Panel C - Plots the false discovery proportion ($FDP$) as a function of the $FDR$. Ideally, the $FDR$ should be equal to the $FDP$. Therefore larger distances from the diagonal are worse.

```{r FDRfdp, fig.cap = "(ref:FDRfdp)", out.width = '90%', fig.width=12, fig.height=10}
dd <- purrr::map_df(allBenchmarks, function(x){res <- x$smc$summary; res$name <- x$model_name;res})
dd <- dd |> dplyr::mutate(nrcontrasts = protein_Id * (4 - nr_missing))
dds <- dd |> dplyr::group_by(name) |> dplyr::summarize(nrcontrasts = sum(nrcontrasts))
dds$percent <- dds$nrcontrasts/max(dds$nrcontrasts) * 100

@@

We can conclude that if we want to sort the proteins according to the likelihood of being differentially regulated to perform gene set enrichment analysis [@subramanian2005gene], the $t$-statistic is better suited than the fold-change estimate. Modeling the degrees of freedom when computing the $p$-values improves the inference (see Figure \@ref(fig:FDRfdp) panel B).
There is no such improvement for the mixed effect model, because the degrees of freedom are erroneously estimated. Furthermore, for the fixed effect linear model, the empirical Bayes variance shrinkage, as suggested by @Smyth2004linear, consistently improves the ranking of proteins compared with the unmoderated estimates (not shown). However, since also for this method, a correct degree of freedom estimate is required, it does not work for mixed effect models.

Computing the statistics at the peptide level, e.g., the $t$-statistics or $p$-value, then summarizing these statistics using their median produces the highest $AUC$ scores among all the tested models (see Figure \@ref(fig:FDRfdp) Panel A left). Furthermore, by using the Beta distribution to model the number of peptides observed, we can further improve the $pAUC$ scores (see Figure \@ref(fig:FDRfdp) Panel A center). However, the properties of Beta-based probabilities are not well understood; for instance, the p-values are not uniformly distributed under the null hypothesis. Furthermore, the $FDR$ estimates obtained when correcting for multiple testing with the Benjamini-Hochberg method are biased and overestimate the false discovery proportion (see Figure \@ref(fig:FDRfdp) Panel C). Therefore, we can not recommend this method if an unbiased estimate of $FDR$ is essential, which is frequently the case. In addition, peptides are stronger affected by missing values, reducing the number of contrasts we could estimate for the dataset using this method (see Figure \@ref(fig:FDRfdp) Panel C).

The R-package `r BiocStyle::Biocpkg("proDA")`, `r BiocStyle::Biocpkg("msqrob2")` and `r BiocStyle::Biocpkg("prolfqua")` (see Methods) model the missing data directly, while MSstats imputes the data using an accelerated failure model. Despite imputation, `r BiocStyle::Biocpkg("MSstats")` estimates group differences for less proteins and does not achieve a higher $pAUC_{10}$. Furthermore, Figure \@ref(fig:FDRfdp) Panel C shows that when using `r BiocStyle::Biocpkg("MSstats")` the proportion of false discoveries might be very high even when filtering for a low $FDR$.

We obtained difference estimates for all proteins, as shown in Figure \@ref(fig:FDRfdp) Panel A when using the probabilistic dropout model (`r BiocStyle::Biocpkg("proDA")`), the hurdle model (`r BiocStyle::Biocpkg("msqrob2")`), and when combining the results of the linear model with contrast estimates in the presence of missing data (prolfqua_merged). Furthermore, the performance of the scaled $p$-values or the $t$-statistics is comparable with that of the other methods (Figure \@ref(fig:FDRfdp) Panel B). All three models perform similarly well when benchmarked against the IonStar/MaxQuant, IonStar/FragPipe, or CPTAC dataset (see Supplement, Benchmark CPTAC). We tested if the $pAUC_{10}$ differences for the three methods were statistically significant, and none of them were at a significance level of $0.4$ (see Supplement, Testing ROC curves). Furthermore, for the IonStar/MaxQuant dataset, msqrob2 performed best, proDA performed best for the CPTAC/MaxQuant data, and prolfqua_merged for the IonStar/FragPipe dataset. Also, the FDR estimates (Figure \@ref(fig:FDRfdp) Panel C) are comparable for all three methods.

>>>>

---
title: prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis
author: 
  - Witold E. Wolski $^{\dagger\ddagger}$^[[Functional Genomics Center Zurich (FGCZ)](https://fgcz.ch) - [University of Zurich](https://uzh.ch)/[ETH Zurich](https://ethz.ch), Winterthurerstrasse 190, CH-8057 Zurich, Switzerland. $^\ddagger$[Swiss Institute of Bioinformatics (SIB), Quartier Sorge - Batiment Amphipole, 1015 Lausanne, Switzerland](https://www.sib.swiss/) $^\dagger$Correspondence&colon; \texttt{wew@fgcz.ethz.ch}]
  - Paolo Nanni$^{\ast}$
  - Jonas Grossmann$^{\ast\ddagger}$
  - Maria d'Errico$^{\ast\ddagger}$
  - Ralph Schlapbach$^{\ast}$
  - Christian Panse$^{\ast\ddagger}$
output: 
  bookdown::pdf_document2:
      toc: true
      toc_appendix: false
      toc_bib: false
      keep_tex: true
      #citation_package: natbib
date: "`r format(Sys.time())`"
bibliography: prolfqua.bib
vignette: >
  %\VignetteIndexEntry{prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis} 
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
abstract: |
    Mass spectrometry is widely used for quantitative proteomics studies, relative protein quantification, and differential expression analysis of proteins. There is a large variety of signal processing software and analysis tools. Nevertheless, there is a need for a modular, easy-to-use application programming interface in R that transparently supports a variety of well principled statistical procedures to make applying them to proteomics data, comparing and understanding their differences easy. The `r BiocStyle::Githubpkg("fgcz/prolfqua")` package integrates essential steps of the mass spectrometry-based differential expression analysis workflow: quality control, data normalization, protein aggregation, statistical modeling, hypothesis testing, and sample size estimation. The package makes integrating new data formats easy. It can be used to model simple experimental designs with a single explanatory variable and complex experiments with multiple factors and hypothesis testing. The implemented methods allow sensitive and specific differential expression analysis. Furthermore, the package implements benchmark functionality that can help to compare data acquisition, data preprocessing, or data modeling methods using a gold standard dataset. The application programmer interface of `r BiocStyle::Githubpkg("fgcz/prolfqua")`  strives to be clear,  predictable, discoverable, and consistent to make proteomics data analysis easy and exciting. Finally, the `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package is available on GitHub https://github.com/fgcz/prolfqua, distributed under the MIT license. It runs on all platforms supported by the [R free software environment for statistical computing and graphics](https://www.r-project.org/).

@@

```{r contrast2, eval = TRUE, echo = TRUE}
mlm <- prolfqua::strategy_lm(
  transformedIntensity ~ condition_ + batch_ + condition_:batch_
  )
Yeast2Factor <- prolfqua::prolfqua_data("data_Yeast2Factor")
mm <- prolfqua::build_model(Yeast2Factor$data, mlm, subject_Id = "protein_Id")

@@

The code snippets in this section demonstrate how a differential expression analysis workflow can be implemented using the `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package. To speed up the computation of these examples, we use a subset of the Ionstar dataset generated by randomly selecting $400$ proteins. First, we remove all proteins with a single peptide and all observations for which MaxQuant reports zero intensities which leaves $332$ proteins. Next, peptide abundances are $\log_2$ transformed and robust z-score scaled using the method `robscale`. Then, using the `LFQDataPlotter` class, we show the distribution of the normalized peptide abundances in Figure \@ref(fig:figure1) Panel A. Afterwards, protein intensities are estimated from peptide intensities using Tukey's median polish. Figure \@ref(fig:figure1) Panel B shows the peptide intensities and the estimated protein intensities. Next, we compute the standard deviation of all the proteins in each group and display their distribution using violin plots (Panel C). Finally, we create a boxplot (Panel D) showing the abundance of one protein.

```{r figure1,  echo = TRUE , fig.cap="(ref:scaling)", out.width = '90%', fig.width = 8, fig.height = 8}
R.version.string; packageVersion("prolfqua")

@@

The following code example illustrates how we compute differences among groups. First, the linear model and the differences are specified. Afterward, the model is fitted to the data using the `build_model` function. Next, we estimate the contrasts either from the linear model using the `Contrasts` class or directly from the data using the `ContrastsSimpleImpute` class. Afterward, we apply $t$-statistic moderation using the `ContrastModerated` class. Finally, the `merge_contrasts_results` function merges both sets of contrast estimates, preferring the one obtained from the linear model if both are available. Then we create the plots shown in Figure \@ref(fig:figure2). Panel A shows the distribution of the $p$-values, Panel B shows the volcano plot for each comparison, and Panel C shows a Bland-Altman plot which shows the difference between groups as a function of the rank of the protein abundance.

(ref:figure2) Panel A - Histogram showing the distribution of p-values for $332$ proteins. Panel B -  Volcano plot showing $-\log_{10}$ transformed $FDR$ as a function of the difference between groups for $332$ proteins. With black dots, we show effect size and FDR estimates obtained from the linear model, while in green, we plot those obtained using imputation. Finally, panel C shows the difference between groups as a function of the rank of the abundance of the proteins.

```{r figure2, fig.cap="(ref:figure2)", echo=TRUE, out.width = '90%', fig.width = 8, fig.height = 5}
# specify differences among groups
contrasts <- c(
  "dilution_(9/7.5)_1.2" =   "dilution.e - dilution.d",
  "dilution_(7.5/6)_1.25" =   "dilution.d - dilution.c"
)
## fit model
lmmodel <- paste(lfqpro$response()," ~ dilution.")
modelFunction <- prolfqua::strategy_lm( lmmodel, model_name = "lm")
models <- prolfqua::build_model(lfqpro, modelFunction)

@@

When comparing differential expression analysis performance, a relevant parameter is the number of proteins for which a method estimated differences (see Figure \@ref(fig:figure3) Panel A), which indicates how robust the procedure works in the presence of missing observations. For each protein, we tried to determine four differences ($\Delta = (1.20, 1.25, 1.30, 1.50)$). Given 4046 proteins with at least two peptides, there are in total $16184$ possible differences. However, some methods can not estimate all of them. The set of proteins with effect size estimates might differ for each method, biasing direct comparison of scores such as the $pAUC$. For instance, we observe that `r BiocStyle::Biocpkg("MSstats")` estimates $16058$ group differences while the mixed effect models estimates the fewest with $15940$. Hence, to conclude that one method shows a better performance, it does not suffice if the $pAUC$ is greater, but the number of proteins with differential expression results needs to be equal or larger.

Figure \@ref(fig:figure3) Panel B shows how various estimates obtained from the models, i.e., the difference between groups, $t$-statistics, and scaled $p$-values allow identifying true positives (TP) given a false positive rate (FPR) of $10%$, by displaying the partial area under the ROC. Ordering proteins by the $t$-statistic or $p$-value leads to a higher $pAUC_{10}$ than when ordering by the estimated difference among groups. Suppose an accurate estimate of the difference among groups is essential. In that case, the linear models fitted to protein intensities, calculated using Tukey's median polish, perform better than those directly modeling peptide intensities, e.g., `ropeca` or `prot_mixed_effect_moderated` (see Table \@ref(tab:describeBenchmarkedModels) Abundance column). We speculate that outliers at the peptide level do not affect the protein estimates when using Tukey's median polish method, a non-parametric method to infer protein abundances. This hypothesis could be examined, by including other forms of protein intensity inferences implemented in `prolfqua`, e.g. `top-N` or `rlm`, into the benchmark.

There are only minor differences in the $pAUC_{10}$ between the $t$-statistics or the scaled $p$-value  (see Figure \@ref(fig:figure3) Panel A). Interestingly, the $pAUC$ increases when using $p$-values instead of the $t$-statistics for linear models, while it decreases for mixed effect models. The reason is an erroneous denominator degree of freedom estimation for many proteins, a known problem in the case of mixed effect models.

We also benchmark if the $FDR$ obtained from a model is an unbiased estimate of the false discovery proportion $FDP$. Figure \@ref(fig:figure3) Panel C shows the $FDP$, obtained from the confusion matrix, as a function of the $FDR$ determined from the model. Most lines are below the diagonal, indicating that the $FDR$ estimates are modestly conservative for this particular benchmark dataset. In the case of `r BiocStyle::Biocpkg("MSstats")`, we observe a high proportion of false discoveries for small $FDR$ values. In the case of the ROPECA method, the $FDR$ estimates, obtained by applying the Benjamini-Hochberg correction to the Beta distribution-based regulation probabilities, strongly overestimate the $FDP$.

(ref:figure3) Panel A - Number of estimated contrasts for each modeling method (higher is better). Panel B - Partial area under the ROC curve at $10\%$ FPR ($pAUC_{10}$) for all contrasts and three different statistics: the difference among groups (diff - Panel B left), the scaled $p$-value (sign(diff) $\cdot$ p.value) (scaled.p.value - Panel B center) and the $t$-statistics (statistic - Panel B right), where a higher pAUC is better. The red line indicates the average area under the curve of all methods.  Panel C - Plots the false discovery proportion ($FDP$) as a function of the $FDR$. Ideally, the $FDR$ should be equal to the $FDP$. Therefore larger distances from the diagonal are worse.

```{r figure3, fig.cap = "(ref:figure3)", out.width = '90%', fig.width=12, fig.height=10}
dd <- purrr::map_df(allBenchmarks, function(x){res <- x$smc$summary; res$name <- x$model_name;res})
dd <- dd |> dplyr::mutate(nrcontrasts = protein_Id * (4 - nr_missing))
dds <- dd |> dplyr::group_by(name) |> dplyr::summarize(nrcontrasts = sum(nrcontrasts))
dds$percent <- dds$nrcontrasts/max(dds$nrcontrasts) * 100

@@

We can conclude that if we want to sort the proteins according to the likelihood of being differentially regulated to perform gene set enrichment analysis [@subramanian2005gene], the $t$-statistic is better suited than the fold-change estimate. Modeling the degrees of freedom when computing the $p$-values improves the inference (see Figure \@ref(fig:figure3) panel B).
There is no such improvement for the mixed effect model, because the degrees of freedom are erroneously estimated. Furthermore, for the fixed effect linear model, the empirical Bayes variance shrinkage, as suggested by @Smyth2004linear, consistently improves the ranking of proteins compared with the unmoderated estimates (not shown). However, since also for this method, a correct degree of freedom estimate is required, it does not work for mixed effect models.

Computing the statistics at the peptide level, e.g., the $t$-statistics or $p$-value, then summarizing these statistics using their median produces the highest $AUC$ scores among all the tested models (see Figure \@ref(fig:figure3) Panel A left). Furthermore, by using the Beta distribution to model the number of peptides observed, we can further improve the $pAUC$ scores (see Figure \@ref(fig:figure3) Panel A center). However, the properties of Beta-based probabilities are not well understood; for instance, the p-values are not uniformly distributed under the null hypothesis. Furthermore, the $FDR$ estimates obtained when correcting for multiple testing with the Benjamini-Hochberg method are biased and overestimate the false discovery proportion (see Figure \@ref(fig:figure3) Panel C). Therefore, we can not recommend this method if an unbiased estimate of $FDR$ is essential, which is frequently the case. In addition, peptides are stronger affected by missing values, reducing the number of contrasts we could estimate for the dataset using this method (see Figure \@ref(fig:figure3) Panel C).

The R-package `r BiocStyle::Biocpkg("proDA")`, `r BiocStyle::Biocpkg("msqrob2")` and `r BiocStyle::Biocpkg("prolfqua")` (see Methods) model the missing data directly, while MSstats imputes the data using an accelerated failure model. Despite imputation, `r BiocStyle::Biocpkg("MSstats")` estimates group differences for less proteins and does not achieve a higher $pAUC_{10}$. Furthermore, Figure \@ref(fig:figure3) Panel C shows that when using `r BiocStyle::Biocpkg("MSstats")` the proportion of false discoveries might be very high even when filtering for a low $FDR$.

We obtained difference estimates for all proteins, as shown in Figure \@ref(fig:figure3) Panel A when using the probabilistic dropout model (`r BiocStyle::Biocpkg("proDA")`), the hurdle model (`r BiocStyle::Biocpkg("msqrob2")`), and when combining the results of the linear model with contrast estimates in the presence of missing data (prolfqua_merged). Furthermore, the performance of the scaled $p$-values or the $t$-statistics is comparable with that of the other methods (Figure \@ref(fig:figure3) Panel B). All three models perform similarly well when benchmarked against the IonStar/MaxQuant, IonStar/FragPipe, or CPTAC dataset (see Supplement, Benchmark CPTAC). We tested if the $pAUC_{10}$ differences for the three methods were statistically significant, and none of them were at a significance level of $0.4$ (see Supplement, Testing ROC curves). Furthermore, for the IonStar/MaxQuant dataset, msqrob2 performed best, proDA performed best for the CPTAC/MaxQuant data, and prolfqua_merged for the IonStar/FragPipe dataset. Also, the FDR estimates (Figure \@ref(fig:figure3) Panel C) are comparable for all three methods.
